Efficient Simulation of Economic and Ecological Dynamics for Marine Spatial Planning

### 2022-10-06 marlin-paper.qmd
revamp with coral experiment

---
title: "Opportunities and precautions for the use of MPAs in bycatch reduction"
author:
    - Daniel Ovando
    - Darcy Bradley
    - Lennon Thomas
    - Echelle Burns
date: "`r Sys.Date()`"
output: 
  bookdown::pdf_document2:
    number_sections: false
  bookdown::html_document2: default
  bookdown::word_document2:
    number_sections: false
    reference_docx: template.docx
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
bibliography: references.bib
params:
  run_name: ["v1.1"]
csl: proceedings-of-the-royal-society-b
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, include = TRUE)

@@

```{r load-results}

@@

<!-- \*\* something to point out: increased catches are a paradox, either -->
<!-- come from rebuilding stock or concentrated fishing, but trajectory can -->
<!-- tell you something. not possible for MPAs to increase catches without -->
<!-- decreasing catches first if they are rebuilding? \*\* -->

<!-- Seasonal movement is kind of interesting, basically just dilutes the -->
<!-- extremes -->

<!-- less resilient species are both more in need of protection and easier to -->
<!-- accidentally fish harder -->

# Abstract

Marine protected areas (MPAs) are a growing part of the marine
conservation toolbox. One of the goals of global MPA expansion is to protect and restore biodiversity through conservation of threatened species often caught incidentally in fisheries, such as marine mammals, turtles, seabirds, and sharks. However, the ability of MPAs to provide population-level conservation benefits to bycatch species depends in part on the ways in which fishing fleets react to MPA placement, and the degree to which habitat and movement dynamics of different species within an ecosystem overlap. In this paper we assess the impacts of MPAs on target and bycatch species and fisheries when faced with fleet redistribution and heterogeneity in life history and habitat use using a novel bio-economic model. We show that while in the majority (xx%) of simulated cases MPAs provided net conservation benefits to all species, in xx% of simulations MPAs resulted in net losses in biomass for 1 or more species. While factors interact in complex and nonlinear ways, placement strategy, adult and larval movement, and degree of habitat heterogeneity were among the most important factors in determining whether MPAs produced a net conservation loss. We then used our model to assess the potential impacts of MPAs in the Western and Central Pacific Ocean on tuna, billfish, and shark populations and fisheries. MPAs produced net conservation gains for all species in nearly all simulations, but did so at the expense of both yields and profits, even when MPA design was optimized. Our results demonstrate the importance of considering fleet behavior, movement dynamics, and habitat heterogeneity in assessing the potential impacts of MPAs in a multi-species and multi-fleet context. XX some punchier closing sentence XX.

# To Do

- Produce results that design MPAs around only a subset of species, to see how sensitive results are to say not including some species in the analysis

- Expand more on the WCPO case study and make the paper a bit more *about* that: make clearer how the model could be used in a real-world setting

- Link better to Will White and company's work on MPA modeling etc

- Include partial dependency plots for variable importance plot

# Introduction

<!-- MAKE IT MORE ABOUT BYCATCH-->

<!-- Empirically, how do you differentiate increase in catch from fishing harder vs increases in catch from rebuilding: rebuilding has to have a decrease at first -->

Marine protected areas (MPAs) are a growing part of the marine
conservation toolbox, with 5.9% of the world's oceans currently in
implemented MPAs [@grorud-colvert2021], and bodies such as the IUCN
calling for 30% protection by 2030. One of the potential benefits of
MPAs is their ability to provide protection to broad segments of an
ecosystem at once, rather than requiring targeted management based on
individual species that may be unfeasible in many parts of the world
lacking in data or fisheries management capacity. Their ability to
provide protection across a range of species with one policy has led them to be suggested as a tool to reduce problems around weak stock or "bycatch"
species [@hastings2017]. We use the term "bycatch" here to generally refer to the
capture of species within a fishery primarily targeting a different set
of species.

While the effect of MPAs on fisheries yields is a continuously debated
topic, the conservation impacts of MPAs are seemingly simpler, albeit
challenging to detect in many cases [@ovando2021a], due in part perhaps
to the perceived obvious nature of their benefits: if you prevent
fishing of a species within an area, there will be more of that species
in that area, and so the conservation benefits of MPAs will increase
with MPA size. @hastings2017 clearly illustrates this idea. Using a two-species two-patch analytical model, they derive conditions
under which MPAs can solve the "weak-stock" problem in which a less
productive species is caught as bycatch by a fleet targeting a more
valuable and productive stock.

However, as @hastings2017 point out, their results depend on a very
specific set of assumptions, namely that 1) adults are perfectly
sedentary 2) the target and bycatch species overlap perfectly in terms
of habitat and 3) recruits are perfectly mixed throughout the system. In
this paper We build on @hastings2017 to consider the role of MPAs in
reducing bycatch while managing tradeoffs with fisheries under a more
resolved model.

Policy leakage is not a unique feature to MPAs, e.g. XX include non-marine examples XX, but
the potential impacts of MPAs on target and bycatch species taking into
account the potential for these leakage effects has received relatively
little generalized attention in the literature [see @abbott2012;
@savina2008; @fulton2011a; @baum2003 for notable exceptions].

We address this gap by providing a strategic evaluation of the sensitivity of bycatch
populations and targeted fisheries to key unknowns and design choices in
MPA planning. We used a novel bio-economic simulation
model that allows for efficient evaluation of the effects of MPAs on
target and bycatch species across a range of ecological and economic
dynamics, critically accounting for the potential impacts of effort
displacement and technical interactions.  We used this model first to
explore general conditions that lead to positive and negative outcomes
for target and bycatch species from MPAs. We then present a case study
tuned to reflect the dynamics of the tuna, billfish, and shark
populations and fisheries of the Western and Central Pacific Ocean
(WCPO) to explore the value of optimal MPA design in managing tradeoffs
between maximizing bycatch conservation and minimizing economic impacts.

We chose the WCPO as a case study as the modern science of MPAs was
largely developed around coastal ecosystems, often characterized by
relatively static features such as coral reefs [citation]. However,
large MPAs are increasingly being implemented and considered across
pelagic ecosystems [@hilborn2021c]. These pelagic systems are characterized by highly
mobile species and fishing fleets using a diverse mix of static and
dynamic ocean features [@game2009; @lubchenco2003]. These ecosystems are
also home to combinations of relatively abundant species targeted by
fishing, such as many tuna stocks [@pons2016], and highly threatened
species caught as incidental "bycatch" in targeted fishing operations,
such as oceanic whitetip sharks [@pacoureau2021]. These dynamics,
combined with the increased policy interest in large MPAs covering
expanses of blue-water ecosystems, make pelagic systems such as the WCPO
an ideal case study in bycatch management through MPAs.

# Methods

@hastings2017 employed an analytically tractable model to derive
conditions under which MPAs would benefit bycatch species. However they
also pointed out that even small increases in model complexity quickly
make analytical solutions to the effects of MPAs intractable. As such,
we employ a numerical simulation strategy in the form of a spatially
explicit, multi-species, multi-fleet, bio-economic model we call
[`marlin`](https://github.com/DanOvando/marlin). Models of MPAs span a
wide range of complexity, from simple analytical tools to end-to-end
ecosystem models [@fulton2015]. `marlin` follows a "models of
intermediate complexity" (MICE) philosophy [@plagányi2014] targeted
specifically around the role of technical interactions across species.
Many MPA models simulate the impacts of MPAs on a single-species level
[@ovando2021a]. This can be useful for for example examining the ways in
which species-specific life history traits interact with fleet dynamics
to influence MPA outcomes. However, bycatch problems are driven by
technical interactions, in which fishing effort for a target species
incidentally captures another species. `marlin` accounts for this by
allowing the user to simulate an arbitrary number of fleets capable of
catching an arbitrary number of species in a spatial area. Different
fleets can place different economic value on different species, while
also having species-specific selectivity functions. This allows us to
simulate bycatch dynamics of MPAs in pelagic ecosystems.

@@

####  Fleet Dynamics

@@

#### Population Model

@@

```{=tex}
\begin{equation}
  ssb_{t,p,s} = \sum_{a=1}^Aw_{a,s,t}^{\gamma_s} mat_{t,a} n_{t,p,s,a}
    (\#eq:ssb)
\end{equation}
```
##### Recruitment {.unnumbered}

@@

```{=tex}
\begin{equation}
  larv_{t,p} = ssb_{t-1} \times\boldsymbol{d^l}
  (\#eq:larvmove)
\end{equation}
```
```{=tex}
\begin{equation}
n_{t,a = 1,p} = \left(\frac{0.8{\times}r0_{p}\times{h}\times{larv_{t,p}}}{0.2\times{ssb0_p}\times(1 - h)+(h - 0.2)\times{larv_{t,p}}}\right) \times \epsilon_t
  (\#eq:dd4)
\end{equation}
```
##### Dispersal {.unnumbered}

Dispersal in the model is broken into two components: adults and
recruits Both assume a Gaussian dispersal kernel of the form

```{=tex}
\begin{equation}
m_{s,p_i,p_j}=\frac{1}{\sqrt{2\pi\sigma_{s,p_i}^2}}e^{-\frac{d_{p_i,p_j}^2}{2\sigma_{s,p_i}^2}}
  (\#eq:move)
\end{equation}
```
where *i* is the source patch, *j* is the destination patch, *d* is the
distance between patches *i* and *j* , and $\sigma_s*$is the movement
rate, in units of patches.

Along with distance, net movement is defined by habitat quality. Habitat
quality defines the quality for adults or larvae of a given patch. The
net probability of movement from patch *i* to patch *j* is defined then
by the product of distance and habitat quality, normalized to sum to one
from each patch, such that all else being equal organisms will move the
most to the closest patch with the best habitat. We allow for seasonal
movement in the model, where years are broken into two seasons. Movement
in season two is defined as the opposite of movement in season one,
intended to reflect the behavior of seasonal migrations.

## Rule-of-thumb MPA Design Strategies

@@

@@

@@

## WCPO Case Study Application

The purpose of the case study is to assess the value of optimization.
Given the time constraints of optimization, rather than iterating over
the `r nrow(state_experiments)`, we fix the values from that table to
values that best reflect the state of
`r n_distinct(experiment_obj$critter)` species reported as caught in the
WCPO.

Sigh. lots of details to add in here. This is the worst.

```{r, echo = FALSE}
knitr::kable(sci_to_com %>%
               rename("Scientific Name" = "critter", "Common Name" = common_name))
```

## Assessing Performance

Each model run started at equilibrium conditions and was then run for
`r years` years (`r seasons * years` seasons. We assessed performance by
comparing relevant metrics (e.g. depletion and profits) under a given
scenario with MPAs relative to the values of those same metrics in the
simulated counterfactual world identical in every way save for the
presence of MPAs.

Degree of habitat difference was calculated as

xx will work out the notation for this xx

$$
 Habitat Difference = sd(sqrt(hab))
$$

In order to weight habitat difference by cells where animals occur, not
where they are not.

# Results

Our simulation experiments allow us to examine the wide range of effects
that MPAs can have on groups of target and bycatch species depending on
combinations of ecological, economic, and design parameters. The goal of
this exercise is not to assign any probability to the potential impacts
of MPAs; we have not provided any explicit probabilities to the
alternative states modeled here, and so it may be that a state that was
only modeled on a few occasions in our simulations is in fact highly
common in the real world.

What these experiments are designed to do is illustrate the range of
outcomes that are possible from MPAs under the types of conditions
evaluated here, and evaluate which drivers are potentially the most
influential in affecting particular outcomes with regards to bycatch.
Our case study of a WCPO-esque system is intended then to reduce the
parameter space enough that we can examine the relative tradeoffs and
value of optimization for economic and conservation objectives in a
pelagic system.

One clear caveat across all of these results is that we have made two
very strong assumptions around the behavior of the fishing fleet. The
first is that we assume that effort is displaced, not removed, by the
creation of MPAs. This is a widely debated question around MPAs [@hilborn2021c], and the
real impacts of MPAs on fishing effort will clearly be setting
dependent. For example, large-scale pelagic fishing fleets already have
high rates of vessel movement built into their economic models, and so
may be more likely to move rather than exit the fishery in response to
MPAs. In contrast, smaller scale coastal fishing operations may simply
exit the fishery if their local fishing grounds are placed in MPAs.
There is clear evidence of effort displacement in many fisheries though
(citation), and moreover a motivation for this study was to evaluate the
performance of MPAs in reducing bycatch given the potential for effort
displacement and technical interactions. If MPAs do not produce
increases in fishing effort outside their borders, then the direct
conservation effects of MPAs have to be positive for all species
concerned.

The second is that we do not account for evolution of total effort. Our
assumption is that while fleets can make economic decisions based around
where to allocate their effort in space, the total pool of effort they
exert is fixed (we would note that the underling `marlin` model does
have capacity for more realistic fleet models such as open-access
dynamics). This is clearly an incorrect assumption over the long term,
but allows us to focus on the impacts of the other drivers considered
here. A more through consideration of the impacts of fleet model
structure on MPA outcomes is generally needed, and is critical for any
tactical efforts at MPA modeling.

<!-- what we found was a series of necessary but not sufficient conditions  -->

<!-- make the tautology clear: illustrating a point, not proving a thing since clearly depends on if you assume displacement, and then degree of non-overlap etc.  -->

We focused first on the range of impacts that MPAs can have on mature
biomass of target and bycatch species across the range of variables
explored. To illustrate, we pulled out two example simulations, one
where MPAs provided relatively large benefits across all three species,
and a second where MPAs actually produced substantial losses in some
species. In the "best case" scenario, each of the three species shares a
common habitat distribution, and so protection of the central habitat of
one species results in protection for all species. In the "worst case"
scenario, an MPA designed primarily around the tuna fishery displaces
effort onto the marlin and shark habitats, increasing the tuna
population while driving the other two species down
(Fig.\@ref(fig:fig-info)).

```{r fig-info, fig.cap="Illustrative example demonstrating how degrees of overlap can influence "}

experiment_summary <- total_experiment_results %>%
  filter(prop_mpa < 0.5,
         !placement_strategy %in% c("area", "avoid_fishing"),
         placement_strategy == "rate",
         placement_error == min(placement_error)) %>%
  group_by(placement_id, state_id) %>%
  summarise(
    max_loss = max(signif_loss),
    min_loss = max(n_loss),
    total_gain = sum(abs_delta_biodiv)
  ) %>%
  ungroup()

worst_case <- experiment_summary %>%
  # filter(max_loss == max(max_loss)) %>%
  arrange(total_gain) %>%
  # filter(total_gain == min(total_gain)) %>%
  slice(2)

worst_state <- state_experiments %>%
  filter(state_id == worst_case$state_id)

worst_result <- compact_experiment_results %>%
  filter(placement_id == worst_case$placement_id,
         state_id == worst_case$state_id,
         prop_mpa <0.5)

worst_result_plot <- worst_result %>%
  ggplot() +
  geom_text(data = tibble(
    x = .1,
    y = max(worst_result$biodiv - worst_result$bau_biodiv) * .9
  ),
  aes(x = x, y = y, label = "Worst Case"),
  size = 4) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_line(aes(prop_mpa, biodiv - bau_biodiv, color = critter),
            show.legend = FALSE, size = 1.25) +
  scale_x_continuous(labels = scales::label_percent(accuracy = 1), name = "MPA Size") +
  scale_y_continuous(
    name = ''
  )

worst_habitat <- worst_state %>%
  select(state_id, data) %>%
  unnest(cols = data) %>%
  select(scientific_name, contains("sigma_"), habitat) %>%
  unnest(cols = habitat) %>%
  mutate(
    scientific_name = fct_recode(
      scientific_name,
      "Shark" = "carcharhinus longimanus",
      "Tuna" = "katsuwonus pelamis",
      "Marlin" = "kajikia audax"
    )
  )

worst_habitat_plot <- worst_habitat %>%
  ggplot(aes(x, y, z = (habitat), color = scientific_name)) +
  geom_contour(show.legend = FALSE) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank(),
    panel.border = element_blank()
  )

best_case <- experiment_summary %>%
  filter(min_loss == min(min_loss, na.rm = TRUE)) %>%
  filter(total_gain == max(total_gain, na.rm = TRUE)) %>%
  slice(1)

best_state <- state_experiments %>%
  filter(state_id == best_case$state_id)

best_result <- compact_experiment_results %>%
  filter(placement_id == best_case$placement_id,
         state_id == best_case$state_id,
         prop_mpa < 0.5)

best_result_plot <- best_result %>%
  ggplot() +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_line(aes(prop_mpa, biodiv - bau_biodiv, color = critter), size = 1.25) +
  geom_text(data = tibble(
    x = .1,
    y = max(best_result$biodiv - best_result$bau_biodiv) * .9
  ),
  aes(x = x, y = y, label = "Best Case"),
  size = 4) +
  scale_x_continuous(labels = scales::label_percent(accuracy = 1), name = "") +
  scale_y_continuous("Change in SSB/SSB0") +
  theme(legend.position = "top") +
  scale_color_discrete(name = element_blank()) +
  labs(title = "A) ")

best_habitat <- best_state %>%
  select(state_id, data) %>%
  unnest(cols = data) %>%
  select(scientific_name, contains("sigma_"), habitat) %>%
  unnest(cols = habitat) %>%
  mutate(
    scientific_name = fct_recode(
      scientific_name,
      "Shark" = "carcharhinus longimanus",
      "Tuna" = "katsuwonus pelamis",
      "Marlin" = "kajikia audax"
    )
  )

best_habitat_plot <- best_habitat %>%
  ggplot(aes(x, y, z = (habitat), color = scientific_name)) +
  geom_contour(show.legend = FALSE) +
  labs(title=  "B)", subtitle = "Habitat Distribution") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank(),
    panel.border = element_blank()
  ) # +
  # facet_wrap(~scientific_name)

fig_info <- (best_result_plot / worst_result_plot) |
  (best_habitat_plot / worst_habitat_plot)

fig_info

```{r}

critter_results <- compact_experiment_results %>%
  filter(bau_biodiv > 0, prop_mpa > 0) %>% 
  group_by(state_id, placement_id) %>%
  mutate(real_sigma_centroid = sd(centroid)) %>%
  mutate(delta_biodiv = pmin(1,biodiv - bau_biodiv),
         loss = biodiv < bau_biodiv,
         real_loss = biodiv < (0.9 * bau_biodiv),
         rec_form = as.factor(rec_form)) %>% 
  mutate(critter = fct_relevel(critter,c("Tuna", "Marlin","Shark")),
         rec_form = fct_recode(rec_form, "Global" = "0", "Local" = "1", "Local Then Disperse" = "2", "Disperse Then Local" = "3"))

Looking at the full
`r n_distinct(total_experiment_results$placement_id) * n_distinct(total_experiment_results$state_id)`
experiments run gains were much more common than losses, with losses of any
magnitude occurring in `r scales::percent(mean(critter_results$loss))`
of simulations with MPAs, but losses of 10% or greater of SSB in only
`r scales::percent(mean(critter_results$real_loss))` of simulations.
However we stress again that we can assign no probabilities to any of
these outcomes in the real world; this simply tells us that under the
range of conditions, selected to represent plausible states of nature,
there are many more ways for MPAs to provide a conservation effect
greater than or equal to 0 than there are for them to provide a net loss
for some species.

The question of interest then are which of the considered drivers appear
to be the most influential in affecting the magnitude and direction of
conservation effects of MPAs. Fishing mortality is the most obvious
answer; if fishing mortality is zero in the absence of an MPA then
clearly a no-take MPA will have no effect. As such we do not focus on
that metric here, focusing instead on other key drivers identified by
our simulation. Four visually prominent factors were MPA size, species,
design strategy, and degree of habitat difference. The degree to which
MPAs of a given size impact spawning biomass depends on a host of
factors. Any potential for conservation losses are intuitively convex in
MPA size: effects of any kind are low when MPA size is low, and
conservation losses must be zero when MPA size is 100% (which this graph
does not quite reach). Peak losses in our simulations occurring
between MPA sizes of XX and XX. Life history plays a clear role our
stylized with shark and marlin populations being much more likely to
experience net conservation losses as a function of MPA placement than
the more resilient tunas. High degree of habitat differences can lead to
both to the potential for large conservation gains or losses depending
on whether a more isolated species is located primarily inside or
outside of MPAs (Fig.\@ref(fig:fig-exp-biodiv)).

```{r fig-exp-biodiv, fig.cap = "Trajectories of changes in SSB/SSB0 relative to BAU for each species across rule-of-thumb MPA design scenarios. Color indicates degree of habitat difference."}

fig_exp_biodiv <- compact_experiment_results %>%
  filter(prop_mpa < 1) %>%
  # filter(f_v_m > 1.5) %>%
  group_by(state_id, placement_id) %>%
  ggplot(aes(
    prop_mpa,
    pmin(1,biodiv - bau_biodiv),
    color =  sigma_habitat,
    group = interaction(placement_id, state_id)
  )) +
  geom_line(alpha = 0.5) +
  geom_hline(aes(yintercept = 0), linetype = 2, color = "red") +
  facet_grid(critter ~ placement_strategy, labeller = labeller(placement_strategy = capitalize)) +
  scale_color_viridis_c(name = "Habitat Difference", guide = guide_colorbar(frame.colour = "black",ticks.colour = "black", barwidth = unit ("15", "lines"))) +
  scale_y_continuous("Change in SSB/SSB0 relative to BAU") +
  scale_x_continuous(labels = scales::label_percent(accuracy = 1), name = "MPA Size") +
  theme(legend.position = "top")

fig_exp_biodiv
```

We can somewhat quantitatively evaluate the relative impact of these
drivers using a regression, given that by construction variation in each
of evaluated drivers are independent in our simulations. We fit a
binomial regression to the presence or absence of of a net conservation
loss per species as a function of the simulated experimental variables.
The factors that according to this model most increase the probability
of a net conservation loss occurring include habitat differences, a
placement strategy of avoiding fishing areas, and the life history of
the species (shark or marlin relative to tuna). The factors that most
decrease the probability of a conservation loss from an MPA are MPA
size, placement strategy based on local depletion, and somewhat
interestingly fishing mortality rate
(Fig.\@ref(fig:fig-loss-drivers)-A).

However, this result is somewhat misleading (xx so should we just
remove?xx). Fig.\@ref(fig:fig-exp-biodiv) shows the clearly bimodal
nature of for example habitat differences. This analysis helps inform us
that habitat differences are a necessary but not sufficient condition
for conservation losses induced by effort displacement. The regression
based results require assuming that each of the effects are independent
and linear in nature. In fact each of these factors can interact in
complex and non-linear ways that may not be apparent visually or through
this type of regression analysis. To address this, we ran a random
forest using the `ranger` package [@wright2017] in R [@rcoreteam2021] and
extracted the impurity corrected importance scores for each of the
variables. While this does not tell us the expected marginal effect of
individual drivers, it does tell us which variables are important if our
goal is to accurately predict the potential for unintended consequences
or magnitude of conservation gains in MPAs. Note here that variables
that intuitively should have an impact but have no visually or
statistically (in a linear model) clear impact on the probability of
experiencing a conservation loss such as the dispersal distances of
adults or larvae have importance scores comparable in magnitude to
habitat differences when considered in the context of nonlinear effects
and interactions with other variables
(Fig.\@ref(fig:fig-loss-drivers)-B).

```{r fig-loss-drivers, fig.cap = "Things"}

loss_reg <-   glm(
  loss ~   f_v_m  + critter + seasonal_movement + adult_movement_sigma + rec_form + recruit_movement_sigma + prop_mpa + sigma_habitat + spatial_allocation + placement_strategy + hyper,
  family = binomial(),
  data = critter_results %>% filter(between(prop_mpa,.2,.4))
)

loss_reg_coefs <- broom::tidy(loss_reg) %>%
  mutate(lower = estimate - 1.96 * std.error,
         upper = estimate + 1.96 * std.error) %>%
  mutate(
    term = str_replace(term, "placement_strategy", "Placement Strategy: "),
    term = str_replace(term, "rec_form", "DD Timing: "),
    term = str_replace(term, "spatial_allocation", "Fleet Allocation: "),
    term = str_replace(term, "hyper", "Hyperallometry present"),
    term = str_replace(term, "prop_mpa", "MPA size"),
    term = str_replace(term, "sigma_habitat", "Habitat differences"),
     term = str_remove_all(term, "critter")
  )

loss_coef_plot <- loss_reg_coefs %>%
  ggplot() +
  geom_hline(aes(yintercept = 0)) +
  geom_pointrange(aes(
    forcats::fct_reorder(term, estimate),
    estimate,
    ymin = lower,
    ymax = upper
  )) +
  scale_y_continuous(name = "Log-odds Ratio Effect") +
  scale_x_discrete(name = '') +
  coord_flip()

tmp_critter <-
  critter_results %>% filter(between(prop_mpa, .2, .4)) %>%
  select(-habitat) %>% 
  mutate(fctr_loss = factor(loss))

loss_rf <-   ranger::ranger(
  loss ~   f_v_m  + critter + seasonal_movement + adult_movement_sigma + rec_form + recruit_movement_sigma + prop_mpa + sigma_habitat + spatial_allocation + placement_strategy + hyper,
  data = tmp_critter,
  importance = "impurity"
)

loss_importance <- data.frame(importance = loss_rf$variable.importance %>% sort()) %>% 
  rownames_to_column(var = "term")

loss_importance_plot <- loss_importance %>% 
  ggplot(aes(forcats::fct_reorder(term, importance), importance)) + 
  geom_col() +
  scale_y_continuous(name = "Variable Importance") +
  scale_x_discrete(name = '') +
  coord_flip()

fig_loss_drivers <- loss_coef_plot + loss_importance_plot

fig_loss_drivers
```

```{r fig-delta-drivers, eval = FALSE}

delta_reg <-   rstanarm::stan_glm(
  delta_biodiv ~   f_v_m  + critter + seasonal_movement + adult_movement_sigma + rec_form + recruit_movement_sigma + prop_mpa + sigma_habitat + spatial_allocation + hyper,
  data = critter_results %>% filter(between(prop_mpa,.2,.4)),
  cores = 4,
  chains = 4
)

tmp_critter <-
  critter_results %>% filter(between(prop_mpa, 0.2, 0.4)) %>%
  select(-habitat) %>% 
  mutate(fctr_loss = factor(loss))
# 
# tmp_rcp <- recipes::recipe(loss ~ ., data = tmp_critter) %>% 
#   themis::step_upsample(fctr_loss, over_ratio = 0.75) %>% 
#   recipes::prep(training = tmp_critter, retain = TRUE) %>% 
#   recipes::juice()

delta_rf <-   ranger::ranger(
  delta_biodiv ~   f_v_m  + critter + seasonal_movement + adult_movement_sigma + rec_form + recruit_movement_sigma + prop_mpa + sigma_habitat + spatial_allocation + placement_strategy + hyper,
  data = tmp_critter,
  importance = "impurity"
)

delta_rf$variable.importance %>% sort()

plot(delta_reg)

## WCPO Case Study

<!-- The generalized experiments help us understand the potential impacts of -->

<!-- the evaluated drivers, but also highlight the challenge in making -->

<!-- definitive statements about the effect of a given driver on the conservation impacts of MPAs, given the complex ways in which all of these factors -->

<!-- interact, even within a relatively simple model such as this one. In -->

<!-- order to help illustrate how addressing the kinds of spatial dynamics -->

<!-- and technical interactions included in our model can affect and benefit -->

<!-- marine spatial planning, we then ran a case study intended to loosely -->

<!-- reflect the dynamics of WCPO tuna, billfish, and shark fisheries and -->

>>>>

---
title: "Efficient Simulation of Economic and Ecological Dynamics for Marine Spatial Planning"
author:
    - Daniel Ovando
    - Darcy Bradley
    - Lennon Thomas
    - Echelle Burns
date: "`r Sys.Date()`"
format:
  pdf:
    toc: true
  html: default
  docx: default
execute:
  echo: false
  message: false
  warning: false
bibliography: references.bib
params:
  params:
  run_name: ["v1.1"]
---

```{r setup}
#| echo: false
#| message: false

@@

@@

need to contrast with SEAPODYM

## Introduction

Communities around the world are increasingly looking to spatial
strategies for managing marine ecosystems. For example, the "30x30"
movement calls for implementation of marine protected areas, a form of
spatial management, across 30% of the world's oceans by 2030. While the
objectives of individual spatial management strategies vary widely,
spatial management strategies are often supported based on their ability
to provide protection to broad swaths of marine ecosystems with one
management action, with potentially fewer costs and complications than
for example assessing, setting, and enforcing catch quotas for
individual species.

However, the design of spatial management strategies faces numerous
challenges. Marine ecosystems are highly dynamic and opaque, and
extractive entities such as fishing fleets respond to changes in marine
ecosystems in highly complex ways. These dynamics post challenges even
for single-species assessment and management strategies [@field2006a],
which are only amplified when we consider management policies
specifically designed around spatial features in the oceans. For
example, policies such as MPAs designed for both conservation and food
security must consider factors such as the optimal size and placement of
a protected area network given a wide array of life history,
exploitation levels, fleet dynamics, and preferred habitats, all of
which may change over time, particularly given the impacts of climate
change.

These challenges are amplified by the fact that our knowledge regarding
the design and implementation of spatial management strategies such as
MPAs is largely based on theoretical studies; the size of protected
areas [@ovando2021a] and time required[@nickols2019] to see substantial
effects from many MPAs means that we have relatively little empirical
evidence as to their large-scale effects under a range of
social-ecological settings [@mcclanahan2021].

Many models have been developed to support the theory and design of
spatial management strategies [@fulton2015]. However, they have tended
to fall into one of either highly complex and tactical models designed
for use in a specific location, or to be highly stylized representations
intended to provide theoretical insights with less relevance for highly
specific design questions. This complex landscape presents a challenge
to stakeholders charged with marine spatial planning exercises lacking
accesible models suited for more tactical applications, and a barrier to
the scientific community, wherein comparisons across spatial management
simulation studies are clouded by discrepancies in underlying model
structures beyond differences in the core phenomenon in question.

In part to help address this challenge, we present here a novel
bio-economic modeling framework called `marlin` that allows for
efficient simulation of spatio-temporal biological and economic
dynamics. `marlin` allows users to simulation the impacts spatial
policies such as MPAs across a range of species targeted by various
fleets across heterogeneous and dynamic seascapes. Here we present the
core methods and functionality of this framework, and demonstrate its
use in two xx case-study applications. In the first, we show how
`marlin` can be used to assess critical drivers for positive outcomes
and unintended consequences of MPAs as a bycatch-mitigation strategy. In
the second, we present a more tactical application, demonstrating how
`marlin` can be used to compare and design MPA networks to achieve
societal objectives.

-   MPAs increasingly part of plan

-   A model based field - hard to do experiments at the time and scale
    needed

-   many models out there, but challenges

    -   highly stylized

    -   highly tactical

-   We present a general tool for modeling spatial dynamics of fished
    populations and fishing fleets, with a focus on MPAs

-   We present two case studies showing the utility of this approach

    -   habitat heterogeneity and bycatch

    -   MPA optimization

-   Basis for improved "what if" testing of MPA scenario

## Methods

Models of MPAs span a wide range of complexity, from simple analytical
tools to end-to-end ecosystem models [@fulton2015]. `marlin` follows a
"models of intermediate complexity" (MICE) philosophy [@plagányi2014]
targeted specifically around the role of technical interactions across
species. Many MPA models simulate the impacts of MPAs on a
single-species level [@ovando2021a]. This can be useful for for example
examining the ways in which species-specific life history traits
interact with fleet dynamics to influence MPA outcomes. However, bycatch
problems are driven by technical interactions, in which fishing effort
for a target species incidentally captures another species. `marlin`
accounts for this by allowing the user to simulate an arbitrary number
of fleets capable of catching an arbitrary number of species in a
spatial area. Different fleets can place different economic value on
different species, while also having species-specific selectivity
functions. This allows us to simulate bycatch dynamics of MPAs in
pelagic ecosystems.

@@

### Fleet Dynamics

@@

### Population Model

@@

```{=tex}
\begin{equation}
  ssb_{t,p,s} = \sum_{a=1}^Aw_{a,s,t}^{\gamma_s} mat_{t,a} n_{t,p,s,a}
    (\#eq:ssb)
\end{equation}
```
#### Recruitment {.unnumbered}

@@

```{=tex}
\begin{equation}
  larv_{t,p} = ssb_{t-1} \times\boldsymbol{d^l}
  (\#eq:larvmove)
\end{equation}
```
```{=tex}
\begin{equation}
n_{t,a = 1,p} = \left(\frac{0.8{\times}r0_{p}\times{h}\times{larv_{t,p}}}{0.2\times{ssb0_p}\times(1 - h)+(h - 0.2)\times{larv_{t,p}}}\right) \times \epsilon_t
  (\#eq:dd4)
\end{equation}
```
#### Dispersal

`marlin`'s movement dynamics are based on a continuous-time Markov
chain, as described in @thorson2021a. Within this framework, the model
allows for movement to be broken down into three components of
*diffusion*, *advection* (follow currents), and *taxis* (active movement
towards say a preferred habitat).

xx movement equations xx

While this approach to movement simulation is not commonly seen in the
marine modeling literature, it has numerous advantages that warrant its
seeming complexity. First, when only diffusion is present, the model
will generalize to the familiar dynamics of a Gaussian dispersal kernel
xx check this xx. However, the time component *t* in equation xx allows
the movement model to be run on arbitrary time steps while preserving
its behavior.

Second, the *advection* component allows users to input useful features
such as the outputs of ROMs models.xxx

Third, the *taxis* model allow for more active habitat choices by
species, allowing us to simulate both preferences of species in space
and time efficiently.

In this way then, `marlin` allows users to run anything from a simple
Gaussian dispersal kernel up to a system governed by species that
passively diffuse around a dynamic thermal range

Lastly, this form has the advantage that the parameters are directly
estimable from real data. So, if provided with say CPUE data and a
tagging study, users can estimates the movement parameters in the same
manner as @thorson2021, and then pass their estimated parameters to
`marlin` for simulation.

### Rule-of-thumb MPA Design Strategies

@@

## Reference Points

A note here WHy no reference points? The problem: Fmsy and MSY are
selectivity dependent, and once we allow for different fleets with
different selectivities that critical can move in space, that movement
affects the net selectivity and fishing mortality on the population.

Bmsy is also affected, since if selectivity is very low, you might need
more biomass of immature individuals relative to less biomass of more
mature individuals; a biomass of X can be more or less reproductively
productive depending on the size / age makeup of that biomass.

SSBmsy is slightly more conserved, as over the long term the key things
is the flow of recruits, so we do provide an option for that

There are many different strategies for estimating reference points
conditional on selectivity and regimes, see references. We leave it to
users to find relevant reference point as required by their methods on
their own, using the sort of MSE examples available in the package.

@@

Two features of marine ecosystems that are often omitted from
theoretical and simulation-based considerations of marine spatial
planning are *technical interactions* and *species on the move*. May
canonical MPA models, informed by planning efforts centered around
relatively static features such as coral reefs, have assumed that
species inhabit a fixed and often uniform (e.g. @hastings2017). In
reality, the habitat of many species are heterogeneous in space, and can
experience both seasonal and directional changes, e.g. climate-change
mediated range shifts.

This reality creates a basic challenge to MPA design, in that given the
need to protect an area less than the total range of a species, the more
complex and dynamics the habitat of the species the more challenging it
will be to identify an area that provides effective protection today and
tomorrow. However, *technical interactions* interact with this spatial
complexity to create another and potentially more difficult challenge.
One of the key questions around spatial management strategies such as
MPAs is to what extent do they *remove* vs *displace* threats.

@@
